Unshuffling Permutations

### 2015-09-15 article.tex
article

>>>>

Given three words $u$, $v$ and $w$, $u$ is said to a
\emph{shuffle} of $v$ and $w$ if it
can be formed by interleaving the letters from $v$ and $w$ in a way
that maintains the left-to-right ordering of the letters from each word
(it is a classical textbook exercice to
design an efficient dynamic programming algorithm that determines
whether $u$ is a shuffle of $v$ and $w$).
A word $u$ is said to be a \emph{square} (w.r.t. the shuffle) if there exists
a word $v$ such that $u$ is a shuffle of $v$ with itself.

This paper is intended to study a natural generalisation of the shuffle
to permutations.
Given permutations $\pi$, $\mu$ and $\nu$, $\pi$ is said to be a shuffle of
$\mu$ and $\nu$ if $\pi$ (viewed as a word) is a shuffle of two words that
are order-isomorphic to $\mu$ and $\nu$.
For example, $\pi$ is the shuffle
of $\mu=$ and $\nu=$ as
$\pi$.
A permutation $\pi$ is said to be a square (w.r.t. the shuffle) if there exists
a permutation $\mu$ such that $\pi$ is a shuffle of $\mu$ with itself.
For example, $\pi = 241356$ is a square as it is the shuffle of $213$ with itself
(\emph{e.g.},
$\pi = \dot{2}\ddot{4}\dot{1}\ddot{3}\ddot{5}\dot{6}$, where
$\dot{i}$ and $\ddot{i}$),
whereas $\pi = 4123$ is not.

Formally, the shuffle of two words $u$ and $v$ of $A^*$ is the language
$u \shuffle v$ consisting of all words
$u_1 v_1 u_2 v_2 \ldots u_k v_k$, where $k \geq 0$ and
the $u_i$ and the $v_i$ are the words  of $A^*$
such that $u = u_1 u_2 \ldots u_k$ and
$v = v_1 v_2 \ldots v_k$.
We first briefly review the key results for words
that arise in our context.
Given words $u$, $v_1$ and $v_2$, it can be tested in
$O\left(|u|^2 / \log(|u|)\right)$ time whether or not
$u \in v_1 \shuffle v_2$~\cite{Leeuwen:Nivat:IPL:1982}.
(To the best of our knowledge, the first
$O(|u|^2)$ time algorithm appeared in~\cite{Mansfield:DAM:1983}.
This algorithm can easily be extended to check in polynomial-time
whether or not a word is the shuffle of any fixed number of given words.)
The shuffle $u \shuffle v$ of words $u$ and $v$ can be computed in
$O\left((|u|+|v|) \; \binom{|u|+|v|}{|u|}\right)$
time~\cite{Spehner:TCS:1986}.
An improvement and generalization has been proposed in
\cite{Allauzen:IGM:2000}, where it is proved that,
given words $u_1, u_2, \ldots, u_k$, the shuffle
$\shuffle_{i=1}^{k} u_i$ can be computed in
$O\left(\binom{|u_1|+|u_2|+\ldots+|u_k|}{|u_1|,|u_2|,\ldots,|u_k|}\right)$ time.
Given words $u, v_1, v_2, \ldots, v_n \in A^*$, it is however \NPC to
decide whether or not
$u \in \shuffle_{i=1}^{k} v_i$.
\cite{Mansfield:DAM:1983,Warmuth:Haussler:JCSS:1984}.
This remains true even if the alphabet has size $3$~\cite{Warmuth:Haussler:JCSS:1984}.
Of particular interest,
it is shown in~\cite{Warmuth:Haussler:JCSS:1984} that this problem remains
\NPC even if all
words $v_1, v_2, \ldots, v_n$ are identical,
thereby proving that,
for two words $u$ and $v$, it is \NPC to decide whether or not $u$
is in the iterated shuffle of $v$.
Again, this remains true even if the alphabet has size $3$.
As for squares, given a word $z$ it is \NPC to determine whether
there exists a word $x$ such that $z$ is the shuffle of $x$ with itself
\cite{Buss:Soltys:2014,Rizzi:Vialette:CSR:2013}.
